Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-04 doc/MS_v2.Rmd
updates from final v2 files

To measure bivalent SC length, two image analysis algorithms were used. The first algorithm estimates the total (summed) SC length across bivalents for individual cells (@wang2019_SC). The second algorithm estimates the SC length of individual bivalents (@peterson2019). Both algorithms apply a ‘skeletonizing’ transformation to synapsed chromosomes that produces a single, pixel-wide ‘trace’ of the bivalent shape. Total SC length per cell was quantified from pachytene cell images [@wang2019_SC]. To reduce algorithmic errors in SC isolation, outliers were visually identified at the mouse level and removed from the data set. Mouse averages were calculated from cell-wide total SC lengths in `r formatC( length(  (Total.SC_DF  %>% filter(species == "M.musculus")  )$Random.Name), format="d", big.mark=",")` -- 3,371 out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") )$Random.Name), format="d", big.mark=",")`-- 4,143 cells with MLH1 counts.

@@

SC length of individual bivalents was quantified in pachytene cell images [@peterson2019]. The DNA CrossOver algorithm [@peterson2019] isolates single, straightened bivalent shapes, returning SC length, location of MLH1 foci, and location of CREST (centromere) foci. The algorithm substantially speeds the accurate measurement of bivalents, but it sometimes interprets overlapping bivalents as single bivalents. In our data set, average proportions of bivalents per cell isolated by the algorithm ranged from 
`r min(algorithm.stat.category$mean_pass.rate)` -- 0.51 ( _molossinus^MSM^_ male) to`r max(algorithm.stat.category$mean_pass.rate)` -- 0.72 ( _musculus^KAZ^_ female). From the total set of pachytene cell images, `r formatC(  length(unique(Curated_BivData$Obj.ID ) ) , format="d", big.mark=",")` -- 10,458 bivalent objects were isolated by the algorithm. Following a manual curation, `r formatC( length(unique(cleanCurated_BivData %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` -- 9,829 single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. The curated single bivalent data supplements our cell wide MLH1 count data with MLH1 foci counts for single bivalents. The proportions of bivalents with the same number of MLH1 foci where compared across strains by a chi-squared test using prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we also considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. This “short bivalent” data set included the four or five shortest bivalents, and excluded the -- X bivalent in oocytes. A total of `r sum(Short.biv_mouse_table$nobs )` -- 678 short bivalents were isolated from `r sum(  (Short.biv_mouse_table  %>% filter(sex=="female") )$ncells)` -- 103 oocytes and `r sum(  (Short.biv_mouse_table  %>% filter(sex=="male") )$ncells)` -- 37 spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A “long bivalent” data set was formed from those bivalents above the 4th quartile in SC lengths per cell. A total of `r sum(Long.biv_mouse_table$n_obs)`  long bivalents were isolated from `r sum(  (Long.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Long.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes.

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` -- 14 wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and two additional species of Mus ( _M. spretus_ and _M. spicilegus_ ). Mean MLH1 foci counts for `r sum(main.table$Nmice)` -- 161 mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` -- 22.0 spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` -- 1,867 spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` -- 18.5 oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` -- 1,409 oocytes).
Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the relationship between MLH1 foci in the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains - _musculus^PWD^_ and _molossinus^MSM^_ - exhibit higher MLH1 focus counts in males. While _musculus^SKIVE^_ has a very low degree of sex differences in MLH1 focus counts. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23 MLH1 foci per cell) of foci (Figure 1C).

To further situate variation in house mouse recombination rate within an evolutionary framework, we fit a series of statistical models including subspecies, strain, and sex, to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` -- 187 mice. To investigate the effects of sex, we limited our analysis to `r length( unique( DF.HetC.MixedModel.HQ$strain) )` strains with sufficient observations from both sexes (Table 1).

We began with a full mixed model (M1; see Materials and Methods), showing that strain (random effect p < 10^{-6} -- p < `r 1/10000`), sex (p = 1.5510^{-8} -- `r MLH1.MMlmer_results$sex.results`), subspecies (p=1.7210^{-4} -- `r MLH1.MMlmer_results$sub.sp`), and subspecies*sex (p = 3.110^{-5} -- `r MLH1.MMlmer_results$interaction.pval` ) each significantly affect MLH1 focus count. After a model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A model with only these two variables (M3) identified two strains with particularly strong effects on MLH1 focus count:  _musculus^MSM^_ (p = 3.9910^{-6} -- p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`) and _domesticus^G^_  (p = 1.0410^{-6} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ). In addition, two strains exhibit strain-by-sex interactions: _molossinus^MSM^_  (p = 1.2610^{-4} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ) and _musculus^PWD^_ (p = 3.8610^{-4} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` ).

We next fit models separately for  `r length(male_mouse.avs_for.analysis$mouse)` -- 192 males and `r length(female_mouse.avs_for.analysis$mouse)` -- 144 females (M4). In the male dataset, three strains affect MLH1 focus count (as observed in Figure 1C): _musculus^PWD^_ (p = 6.3110^{-8} -- `r sum_MLH1.male.glm.M4$coefficients[,4][8]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[8]` -- 6.1 foci), and _musculus^SKIVE^_ (p = `r sum_MLH1.male.glm.M4$coefficients[,4][10]` ; effect = 3.8-- `r sum_MLH1.male.glm.M4$coefficients[10]` ), and _molossinus^MSM^_  (p=2.4210^{-12} -- `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]` -- 7.0).

These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. Analysis of the female dataset identifies four strains with effects on recombination rate: _domesticus^G^_ (p = 2.510^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`), _molossinus^MSM^_  (p = 6.2410^{-6} -- `r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), _domesticus^LEW^_ (p = 0.01 -- `r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]` ), and _musculus^PWD^_ (p= 0.02 -- `r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]` ). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males. 
Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

@@

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p=`r Q1.lmer_totSC_M1_results$sex.results`; M2: p=`r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][9,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and short bivalent measures (M1: p= `r Q1.lmer_short_M1_results$sex.results`; M2: p= `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,2] / 10000`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2] / 10000`). The existence of some subspecies and strain effects (M1 interaction: p = `r Q1.lmer_short_M1_results$interaction.pval`; M2: strain-SKIVE `r summ.Q1.glm_short.SC_M2$coefficients[7,c(1,4)][2]`,  SKIVE * male p= `r summ.Q1.glm_short.SC_M2$coefficients[13,c(1,4)][2]`) further indicates that SC length has evolved among strains and among subspecies. In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males have more MLH1 foci per cell. This pattern further suggests that in high-recombination strains, spermatocytes have less space in which to position additional crossovers than do oocytes.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: 0.68 -- `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )` -- 0.56; t-test; p = 2.9210^{-22} -- `r Q1_1CO.pos_t.test.FULL$p.value` ). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = 1.2610^{-25} -- `r lmer.Q1.nrm1CO_M1_results$sex.results` ; M2: p = 1.3310^{-7} -- `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p = 1.3310^{-7} -- `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 1.4910^{-12} -- p < 0.05), with only _musculus^KAZ^_ showing no difference (t-test; p = 0.33 -- `r Q1.IFD.PER_t.test_KAZ$p.value` ). Examination of IFD~norm~ distributions indicates that female values are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Analysis of models treating IFD~norm~ as the dependent variable supports the inference of stronger interference in males, with sex being the most significant variable (M1: p = 6.7410^{-14} --`r Q1.IFD.PER_M1_lmer.results$sex.results` ; M2: p = 0.01 -- `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`, M3: p = 0.01 -- `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`).

@@

```{r algorithm_performance_stats, echo=FALSE}

@@

```{r Chrm.Prop.misc.calulations, echo=FALSE, eval=FALSE}

total.biv.all <- sum( MELT_Chrm.Proportion$n0CO, MELT_Chrm.Proportion$n1CO, MELT_Chrm.Proportion$n2CO, MELT_Chrm.Proportion$n3CO, MELT_Chrm.Proportion$n4CO)

```{r Q2.SC_class.rec.group_setup, echo=FALSE, include=FALSE, eval=FALSE}
#REMOVE THIS, IT uses total bivalent data

#rerun the models for datasets -- deviding by chrm class and rec group
Curated_BivData$Rec.group <- ifelse(grepl("PWD male", Curated_BivData$category), 1, 
                            ifelse(grepl("MSM male", Curated_BivData$category), 1,
                             ifelse(grepl("SKIVE male", Curated_BivData$category), 1, 0)) )

Q2.SC_BivData_low.rec <- Curated_BivData %>% filter(sex=="male") %>% filter( hand.foci.count < 5)%>% filter(Rec.group == 0 )
Q2.SC_BivData_high.rec <- Curated_BivData %>% filter(sex=="male") %>% filter( hand.foci.count < 5) %>% filter(Rec.group == 1 )

#calq_mouse av
mouse.av.SC.low.rec <- ddply(Q2.SC_BivData_low.rec, c("mouse"), summarise,
                        n_obs = length(Obj.ID),
                        n_cells = length(unique(fileName) ),
                        mean.SC.total = mean(chromosomeLength),

                    mean.SC_0CO = mean(chromosomeLength[which(hand.foci.count == 0)], na.rm = TRUE),
                  n_0CO =  sum((hand.foci.count == 0), na.rm = TRUE ),

                   mean.SC_1CO = mean(chromosomeLength[which(hand.foci.count == 1)], na.rm = TRUE),
                    n_1CO = sum((hand.foci.count == 1), na.rm = TRUE),

                   mean.SC_2CO = mean(chromosomeLength[which(hand.foci.count == 2)], na.rm = TRUE),
                   n_2CO = sum((hand.foci.count == 2), na.rm = TRUE),

                  mean.SC_3CO = mean(chromosomeLength[which(hand.foci.count == 3)], na.rm = TRUE),
                  n_3CO = sum((hand.foci.count == 3), na.rm = TRUE)

                         )

mouse.av.SC.low.rec <- add_strain(mouse.av.SC.low.rec)
mouse.av.SC.low.rec <- add_subsp(mouse.av.SC.low.rec)

mouse.av.SC.low.rec$strain <- factor(mouse.av.SC.low.rec$strain, ordered = FALSE)
mouse.av.SC.low.rec$subsp <- factor(mouse.av.SC.low.rec$subsp, ordered = FALSE)

#calq_mouse av
mouse.av.SC.high.rec <- ddply(Q2.SC_BivData_high.rec, c("mouse"), summarise,
                        n_obs = length(Obj.ID),
                         mean.SC = mean(chromosomeLength)
                         )
mouse.av.SC.high.rec <- add_mouse(mouse.av.SC.high.rec)
mouse.av.SC.high.rec <- add_strain(mouse.av.SC.high.rec)
mouse.av.SC.high.rec <- add_subsp(mouse.av.SC.high.rec)

mouse.av.SC.high.rec$strain <- factor(mouse.av.SC.high.rec$strain, ordered = FALSE)
mouse.av.SC.high.rec$add_subsp <- factor(mouse.av.SC.high.rec$subsp, ordered = FALSE)
```

We used the contrast between males from high-recombination strains and males from low-recombination strains to identify features of the recombination landscape associated with evolutionary transitions in the genome-wide recombination rate. We considered proportions of bivalents with different numbers of crossovers, double-strand breaks, SC length, and crossover positions, including crossover interference.

Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")` -- 34,982) have either one or two MLH foci (Figure 3). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Figure 3).

High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` -- 0.33 in _musculus^SKIVE^_, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` -- 0.44 in _musculus^PWD^_ , and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` -- 0.53 in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _musculus^KAZ^_: p = 3.1510^{-33} -- `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = 4.7210^{-13} -- `r MSM.MOLF.2CO.prop.test$p.value`) and _musculus^PWD^_ vs. _molossinus^MSM^_: p = `r PWD.MSM.2CO.prop.test$p.value`. These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number on multiple bivalents.

>>>>

To measure bivalent SC length, two image analysis algorithms were used. The first algorithm estimates the total (summed) SC length across bivalents for individual cells (@wang2019_SC). The second algorithm estimates the SC length of individual bivalents (@peterson2019). Both algorithms apply a ‘skeletonizing’ transformation to synapsed chromosomes that produces a single, pixel-wide ‘trace’ of the bivalent shape. Total SC length per cell was quantified from pachytene cell images [@wang2019_SC]. To reduce algorithmic errors in SC isolation, outliers were visually identified at the mouse level and removed from the data set. Mouse averages were calculated from cell-wide total SC lengths in `r formatC( length(  (Total.SC_DF  %>% filter(species == "M.musculus")  )$Random.Name), format="d", big.mark=",")` out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") )$Random.Name), format="d", big.mark=",")` cells with MLH1 counts.

@@

SC length of individual bivalents was quantified in pachytene cell images [@peterson2019]. The DNA CrossOver algorithm [@peterson2019] isolates single, straightened bivalent shapes, returning SC length, location of MLH1 foci, and location of CREST (centromere) foci. The algorithm substantially speeds the accurate measurement of bivalents, but it sometimes interprets overlapping bivalents as single bivalents. In our data set, average proportions of bivalents per cell isolated by the algorithm ranged from 
`r min(algorithm.stat.category$mean_pass.rate)` ( _molossinus^MSM^_ male) to `r max(algorithm.stat.category$mean_pass.rate)` ( _musculus^KAZ^_ female). From the total set of pachytene cell images, `r formatC(  length(unique(Curated_BivData$Obj.ID ) ) , format="d", big.mark=",")` bivalent objects were isolated by the algorithm. Following a manual curation, `r formatC( length(unique(cleanCurated_BivData %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. The curated single bivalent data supplements our cell wide MLH1 count data with MLH1 foci counts for single bivalents. The proportions of bivalents with the same number of MLH1 foci where compared across strains by a chi-squared test using prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we also considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. This “short bivalent” data set included the four or five shortest bivalents, and excluded the -- X bivalent in oocytes. A total of `r sum(Short.biv_mouse_table$n_obs )` short bivalents were isolated from `r sum(  (Short.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Short.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A “long bivalent” data set was formed from those bivalents above the 4th quartile in SC lengths per cell. A total of `r sum(Long.biv_mouse_table$n_obs)`  long bivalents were isolated from `r sum(  (Long.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Long.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes.

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and two additional species of Mus ( _M. spretus_ and _M. spicilegus_ ). Mean MLH1 foci counts for `r sum(main.table$Nmice)` mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes).

Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the relationship between MLH1 foci in the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains - _musculus^PWD^_ and _molossinus^MSM^_ - exhibit higher MLH1 focus counts in males. While _musculus^SKIVE^_ has a very low degree of sex differences in MLH1 focus counts. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23 MLH1 foci per cell) of foci (Figure 1C).

To further situate variation in house mouse recombination rate within an evolutionary framework, we fit a series of statistical models including subspecies, strain, and sex, to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` mice. To investigate the effects of sex, we limited our analysis to `r length( unique( DF.HetC.MixedModel.HQ$strain) )` strains with sufficient observations from both sexes (Table 1).

We began with a full mixed model (M1; see Materials and Methods), showing that strain (random effect p < `r 1/10000`), sex (p = `r MLH1.MMlmer_results$sex.results`), subspecies (p= `r MLH1.MMlmer_results$sub.sp`), and subspecies*sex (p = `r MLH1.MMlmer_results$interaction.pval` ) each significantly affect MLH1 focus count. After a model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A model with only these two variables (M3) identified two strains with particularly strong effects on MLH1 focus count:  _musculus^MSM^_ (p = 3.9910^{-6} -- p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`) and _domesticus^G^_  (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ). In addition, two strains exhibit strain-by-sex interactions: _molossinus^MSM^_  (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ) and _musculus^PWD^_ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` ).

We next fit models separately for  `r length(male_mouse.avs_for.analysis$mouse)` males and `r length(female_mouse.avs_for.analysis$mouse)` females (M4). In the male dataset, three strains affect MLH1 focus count (as observed in Figure 1C): _musculus^PWD^_ (p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[8]` foci), and _musculus^SKIVE^_ (p = `r sum_MLH1.male.glm.M4$coefficients[,4][10]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[10]` ), and _molossinus^MSM^_  (p = `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]`).

These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. Analysis of the female dataset identifies four strains with effects on recombination rate: _domesticus^G^_ (p = `r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`), _molossinus^MSM^_  (p = `r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), _domesticus^LEW^_ (p = `r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]` ), and _musculus^PWD^_ (p = `r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]` ). Strain effect sizes in females are modest in magnitude (ranging from 1 to 4 foci) compared to those in males. Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

@@

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p = `r Q1.lmer_totSC_M1_results$sex.results`; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][9,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and short bivalent measures (M1: p = `r Q1.lmer_short_M1_results$sex.results`; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,2] / 10000`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2] / 10000`). The existence of some subspecies and strain effects (M1 interaction: p = `r Q1.lmer_short_M1_results$interaction.pval`; M2: strain-SKIVE `r summ.Q1.glm_short.SC_M2$coefficients[7,c(1,4)][2]`,  SKIVE * male p = `r summ.Q1.glm_short.SC_M2$coefficients[13,c(1,4)][2]`) further indicates that SC length has evolved among strains and among subspecies. In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males have more MLH1 foci per cell. This pattern further suggests that in high-recombination strains, spermatocytes have less space in which to position additional crossovers than do oocytes.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; t-test; p = `r Q1_1CO.pos_t.test.FULL$p.value` ). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results` ; M2: p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (all t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (t-test; p = `r Q1.IFD.PER_t.test_KAZ$p.value` ). Examination of IFD~norm~ distributions indicates that female values are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Analysis of models treating IFD~norm~ as the dependent variable supports the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results` ; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`, M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`).

@@

```{r algorithm_performance_stats, echo=FALSE, eval=FALSE}

@@

We used the contrast between males from high-recombination strains and males from low-recombination strains to identify features of the recombination landscape associated with evolutionary transitions in the genome-wide recombination rate. We considered proportions of bivalents with different numbers of crossovers, double-strand breaks, SC length, and crossover positions, including crossover interference. Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")`) have either one or two MLH focus (Figure 3). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Figure 3).

High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` in _musculus^SKIVE^_, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in _musculus^PWD^_ , and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _musculus^KAZ^_: p = `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = `r MSM.MOLF.2CO.prop.test$p.value`) and _musculus^PWD^_ vs. _molossinus^MSM^_: p = `r PWD.MSM.2CO.prop.test$p.value`. These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number on multiple bivalents.

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )`  early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from three low-recombination strains ( _musculus^KAZ^_, _domesticus^WSB^_ , and  _domesticus^G^_) and two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test;  p < `r 1/1000000` ). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval` ). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision.

Although there is a positive correlation between total SC length and MLH1 foci when both traits are measured as mouse means (Spearman’s r = `r cor( mouse.total.skel$mean.skel, mouse.total.skel$mean.MLH1, method = c("spearman") )`; p = `r cor.test(mouse.total.skel$mean.skel, mouse.total.skel$mean.MLH1, method=c("spearman") )$p.value` ), total SC length only partially differentiates high-recombination strains from low-recombination strains (Figure 5). Whereas high-recombination strains as a group have significantly greater total SC length than low-recombination strains (t-test; p = `r Q2tot.sc_t.test_High.Low$p.value`), separate tests within subspecies show that the two strain categories differ within _M. m. molossinus_ (p = `r Q2tot.sc_t.test_High.Low_Mol$p.value` ) but not within _M. m. musculus_ (p = `r Q2tot.sc_t.test_High.Low_Musc$p.value`).

Additionally, mouse means for the reduced (short and long) bivalent datasets do not differ between high-recombination and low-recombination strains (t-test; short: p = `r Q2short.biv_t.test_High.Low$p.value` ; long: p = `r Q2long.biv_t.test_High.Low$p.value`). In a model with total SC length as the dependent variable, two subspecies effects are significant ( _M. m. musculus_ p = `r sum.Q2.glm_mus.av_totSC_M4[2,4]` , _M. m. molossinus_, p = `r sum.Q2.glm_mus.av_totSC_M4[3,][4]` ). In models with SC lengths of short and long bivalents as dependent variables, several subspecies and strain effects reach significance (p < 0.05), but they are not consistent across models. Collectively, these observations reveal that evolution of the genome-wide recombination rate is not strongly associated with evolution of SC length.

High-recombination and low-recombination strains do not differ in MLH1 focus positions on bivalents with a single focus, either for the full dataset (t-test; p = `r Q2.1COpos_Ttest$p.value` ) or within subspecies (t-test; _M. m. musculus_: p = `r Q2.1COpos_mus.av_Musc_ttest$p.value` ; _M. m. molossinus_ : p = `r Q2.1COpos_mus.av_Mol_ttest$p.value` ). While _domesticus^WSB^_ and _molossinus^MOLF^_ exhibit strain effects in a model with normalized crossover position as the dependent variable, these two strains do not differ in MLH1 count. These results show that evolution of crossover position on chromosomes with single crossovers is decoupled from evolution of the genome-wide recombination rate, at least at this scale of resolution.

High-recombination strains have greater inter-focal distances (in both raw and normalized measured) than low-recombination strains (t-test; IFD~norm~: p = `r t.test_mouse.av.IFD_PER$p.value` ; IFD~raw~: p = `r t.test_mouse.av.IFD_ABS$p.value`). This pattern holds within _M. m. musculus_ (t-test; IFD~norm~: p = `r  t.test_mouse.av.IFD_PER_MUSC$p.value`; IFD~raw~: p = `r  t.test_mouse.av.IFD_ABS_MUSC$p.value`), with a trend in the same direction within _M. m. molossinus_ (IFD~norm~: p= `r t.test_mouse.av.IFD_PER_MOL$p.value` ; IFD~raw~: p = `r t.test_mouse.av.IFD_ABS_MOL$p.value`).
